---
title: "Storm Data Analysis"
author: "Bert Lievrouw"
date: "5/6/2018"
output: html_document
---

## Synopsis
This study analyses the effect of weather conditions on population health and the economy in the US based on data gathered from 1950 until november 2011 by the National Weather Service. Additional information about the underlying data can be obtained. [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) and [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)


## Data processing

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Don't use scientific notation
options(scipen = 999)

# Load required packages
library(dplyr)
library(ggplot2)
library(reshape2)
```

The analysis required the R packages `dplyr`, `ggplot2` and `reshape2`.

```{r}
zipfile_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
zipfile_path <- "data/noaa_storm_data.csv.bz2"
```

Data for this analysis was downloaded as a bz2 zipped file from the [web](`r zipfile_url`). The compressed raw csv file with headers was read into an R dataframe. Empty strings were interpreted as NA values.

```{r download_and_read_data, include=TRUE, cache=TRUE}
# Download bz2 zipped file to data directory and read it
if (!dir.exists("data")) {
    dir.create("data")
}
if (!file.exists(zipfile_path)) {
    download.file(zipfile_url, zipfile_path)
}
if (!exists("storm_df")) {
    storm_df <- read.csv(zipfile_path, header=TRUE, na.strings = "")
}
```


Data was grouped by weather event (`EVTYPE`) in order to analyse the health and economic effects of these events.

```{r}
storm_data_by_event <- storm_df %>% group_by(EVTYPE)
```


The impact on population health by weather event type was estimated using the sum of all reported fatality (`FATALITIES`) and injury (`INJURIES`) numbers for these events. A plot was created which reports these totals for the most lethal (i.e. highest fatality totals) weather events.

```{r population_health_analysis}
# Rank total fatalities by event type
fatalities_by_event <- storm_data_by_event %>%
    summarize(total_fatalities=sum(FATALITIES)) %>%
    # Rank events by total fatalities
    arrange(desc(total_fatalities)) %>%
    as.data.frame

# Top 20 total fatalities during recorded period by weather event
head(fatalities_by_event, 20)

# Rank total injuries by event type
injuries_by_event <- storm_data_by_event %>%
    summarize(total_injuries=sum(INJURIES)) %>%
    # Rank events by total injuries
    arrange(desc(total_injuries)) %>%
    as.data.frame

# Top 20 total injuries during recorded period by weather event:
head(injuries_by_event, 20)

# Merge fatality and injury data, primarily sort by fatalities
population_hazards_by_event <- merge(fatalities_by_event, injuries_by_event, by="EVTYPE") %>%
    arrange(desc(total_fatalities), desc(total_injuries))

# Create barplot displaying health effects of 15 most lethal weather events
population_hazards_by_event_head_melt <- melt(population_hazards_by_event[1:15,], id="EVTYPE")
most_fatal_events_ranked <- unique(population_hazards_by_event_head_melt$EVTYPE)
most_lethal_weather_events_plot <- ggplot(population_hazards_by_event_head_melt, 
    aes(x=factor(EVTYPE, levels=most_fatal_events_ranked), y=value, fill=variable)) + 
    geom_bar(stat="identity", position="dodge") + 
    labs(title="Most lethal weather events in US (1950 - 2011)", x="Weather event", y="Population") + 
    scale_fill_discrete(name="Health effect", 
                        breaks=c("total_fatalities", "total_injuries"), 
                        labels=c("Total fatalities", "Total injuries")) + 
    scale_y_log10() + 
    theme(axis.text.x = element_text(angle = 65, hjust = 1))
```


The economic damages by weather event type were estimated using the sum of all reported property damages (`PROPDMG`) and damages to crops (`CROPDMG`).

```{r economic_consequences_analysis}
# Rank total property damages by weather event type
property_damage_by_event <- storm_data_by_event %>%
    summarize(total_property_damage = sum(PROPDMG)) %>%
    arrange(desc(total_property_damage)) %>%
    as.data.frame

# Ranking of weather events with highest total property damages ($) during recorded period
head(property_damage_by_event, 20)

# Rank total crop damages by weather event type
crop_damage_by_event <- storm_data_by_event %>%
    summarize(total_crop_damage = sum(CROPDMG)) %>%
    arrange(desc(total_crop_damage)) %>%
    as.data.frame

# Ranking of weather events with highest total crop damages ($) during recorded period:
head(crop_damage_by_event, 20)

# Merge total property and crop damages and calculate total economic damages by weather event
economic_damage_by_event <- merge(property_damage_by_event, crop_damage_by_event, by="EVTYPE") %>%
    mutate(total_economic_damage = total_property_damage + total_crop_damage) %>%
    arrange(desc(total_economic_damage)) %>%
    as.data.frame

# Weather events with highest total economic damages ($) (property and crop damage)
head(economic_damage_by_event[, c("EVTYPE", "total_economic_damage")], 20)

# Create barplot displaying total economic damage of weather events with highest economic impact
highest_economic_damages_by_weather_events_plot <- ggplot(economic_damage_by_event[1:15,], 
    aes(x=factor(EVTYPE, levels=economic_damage_by_event[1:15,]$EVTYPE), y=total_economic_damage/(10^6))) + 
    geom_bar(stat="identity") + 
    labs(title="Weather events with most economic damages in US (1950 - 2011)", x="Weather event", y="Damages ($M)") + 
    theme(axis.text.x = element_text(angle = 65, hjust = 1))

# Create barplot displaying total economic damage of weather events with highest economic impact as fraction of overall weather related damages
total_economic_damage_overall <- sum(economic_damage_by_event$total_economic_damage)
highest_economic_damages_fraction_by_weather_events_plot <- ggplot(economic_damage_by_event[1:30,], 
    aes(x=factor(EVTYPE, levels=economic_damage_by_event[1:30,]$EVTYPE), y=(total_economic_damage/total_economic_damage_overall)*100)) + 
    geom_bar(stat="identity") + 
    labs(title="Weather events with most economic damages in US (1950 - 2011)", x="Weather event", y="Damages (%)") + 
    theme(axis.text.x = element_text(angle = 65, hjust = 1))
```


## Results

### Population health

- Total **fatalities and injuries** of the most lethal weather events:
```{r}
most_lethal_weather_events_plot
```

Tornadoes seem to have been by far the most harmful for the health of the US population from 1950 until november 2011.



### Economic consequences

- Ranking of weather events with highest **total economic damages** (property + crop damage) during recorded period:
```{r}
highest_economic_damages_by_weather_events_plot
highest_economic_damages_fraction_by_weather_events_plot
```

More then a quarter of all weather related economic damages in the US from 1950 until november 2011 are attributable to tornadoes.

```{r}
head(crop_damage_by_event, 3)
```

Hail is the biggest contributor to crop related economic damages



